Trends in the contribution of major causes of death to the black-white life expectancy gap by US state

### 2018-05-10 Manuscript/Manuscript_COD_Trends_HealthandPlace.Rmd
updated methods

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEERStat) to extract data on the number of deaths and the population size in 39 US states and the District of Columbia between 1969 and 2013.^5^ We excluded 11 states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 0.4% of the black population in the United States in 2010.^6^ The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System.^7^ The population data were derived by the NCHS using intercensal bridged-race population estimates. We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 to 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes (**Table S1**). We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census.^8^

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppresses counts of deaths between 1 and 9,^9^ which we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

We used a Bayesian time-series model to smooth the death counts across years and estimate the underlying mortality rate. Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than 9 deaths. The mortality rates over time within each stratum were assigned an autoregressive prior, where the log mortality rate was assumed to be Gaussian centered at the previous year's mortality rate. A non-informative gamma prior was used to estimate the inverse variance. For censored strata with between one and nine deaths, the deaths were modeled using a truncated Poisson likelihood, and otherwise the model remained the same. The truncated Poisson was bound between one and the larger of nine or the stratum's population size, which could be less than nine.

Models were fit using Markov chain Monte Carlo (MCMC). These models were run separately for every combination of state, race, gender, and cause of death. They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence. One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples. Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles. This modeling was done in R (version 3.3.2) using the *rjags* package.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total number of deaths in each age group, and standard life table methods were applied.^10,11^ Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method.^12^ This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to cardiovascular disease among blacks compared to whites. If the decomposition produces a negative estimate this implies that whites have a higher rate of mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing 40 state-level trends and their credible intervals.

We also calculated state-level trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000.^13^ The calculations described in this section were performed using functions written by the first author.

@@

A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on Github: link [DOI: removed for double blind review]

>>>>

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER\*Stat) to extract data on the number of deaths and the population size in thirty-nine US states and the District of Columbia (DC) between 1969 and 2013 (National Cancer Institute, 2016). We excluded eleven states with the smallest black populations: Alaska, Hawaii, Idaho, Maine, Montana, New Hampshire, North Dakota, South Dakota, Utah, Vermont, and Wyoming. These states contained less than 10,000 black residents in the first years' of data, which led to computational issues when we tested our models. Overall, these states comprised less 0.4% of the black population in the United States in 2010, implying that their removal will not significantly detract from the depiction of racial inequality in the US that we aim to provide (U.S. Census Bureau, 2017a). The mortality data were originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System (Centers for Disease Control and Prevention, 2017). A thorough description of the population denominators provided by SEER\*Stat is available online (National Cancer Institute, 2018). Briefly, the population denominators were derived from US Census data. Prior to 2000, US residents could only check one racial categorization (American Indian or Alaskan Native, Asian or Pacific Islander, Black, or White). In 2000, multiple racial categories could be checked on the Census but the four-race categorization is still in-use by vital statistics. Thus, the population denominators in these years were derived by the NCHS using intercensal bridged-race estimates (2000-2009) or postcensal bridged race estimates (2010-2013), which used algorithms to go from the multiple race categorizations to the four-race system (Ingram et al., 2003). We used data on the number of deaths and the population size according to race (black or white), gender (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969-2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were aggregated into six broad causes of death for this analysis (Supplementary Appendix 1 Table S1). Major categories are: Cardiovascular disease (the most common cause of death), Cancers (the second most common cause of death), Non-communicable disease (often related to diabetes, obesity, or aging, three risk factors that became more prevalent over the study time period), Communicable disease (capturing periods of increased infectious activity), Injuries (capturing external causes of death, and suicide, which are of increased importance in early adulthood), and All other causes (capturing infant mortality, maternal mortality, and ill-defined causes). We chose these six categories to balance the limitations of parsing the data too finely, which would lead to data suppression and grouping causes too broadly, which would limit their interpretability. These major causes also approximate what has been used in previous research and aid extension of these earlier findings (Harper et al., 2007; Harper et al., 2012). We did not consider Hispanic ethnicity as part of our racial definition, since Hispanic origin was ascertained beginning in 1978 on death certificates in certain states and 1980 for the Census (Arial et al., 2008).

@@

The data structure produced two statistical challenges that were addressed by our modeling strategy. Firstly, because of confidentiality, NCHS suppressed counts of deaths between one and nine, (National Cancer Institute, 2017a) which we needed to impute. Secondly, several states had small black populations which produced sparse counts in some state-year-race-sex-age-cause strata. In these cases, the estimation of the mortality rates is imprecise, and can produce abrupt changes in the estimated mortality rates over time that are unrealistic. To better estimate the underlying mortality rates, we used a method that produces smoother age-specific mortality trends.

@@

We used a Bayesian time-series model to smooth the death counts across years and estimate the underlying mortality rate. Specifically, we used a Bayesian generalized linear model with a Poisson likelihood to model the number of deaths in each state-year-sex-race-age-cause stratum as a function of the at-risk population, for strata with zero or more than nine deaths. The mortality rates over time within each stratum were assigned an autoregressive prior, where the log mortality rate was assumed to be Gaussian centered at the previous year's mortality rate. A non-informative gamma prior was used to estimate the inverse variance. For censored strata with between one and nine deaths, the deaths were modeled using a truncated Poisson likelihood, and otherwise the model remained the same. The truncated Poisson was bound between one and the larger of nine or the stratum's population size, which could be less than nine.

Models were fit using Markov chain Monte Carlo (MCMC). These models were run separately for every combination of state, race, gender, and cause of death. They were fit using two chains of length 10,000, with a 2,000 iteration burn-in period to ensure convergence. One thousand iterations remained after burn-in and sampling every eighth iteration to reduce autocorrelation among the MCMC samples. Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions formed the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles. This modeling was done in R (version 3.3.2) using the *rjags* package.

@@

To calculate life expectancy at birth, the smoothed cause-specific death counts were aggregated to estimate the total number of deaths in each age group, and standard life table methods were applied (Preston et al., 2001; Selvin, 2004). Life expectancy at birth (in years) was calculated for blacks and whites in each state and year, by gender. The difference in life expectancy between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in each state, year, and gender.

To examine the components of the difference in life expectancy between blacks and whites, we performed a decomposition analysis by cause of death using the Arriaga method (Arriaga, 1985). This method partitions the black-white difference in life expectancy by cause of death, which can be interpreted as the amount of the gap, in years, that can be attributed to higher cause-specific mortality rates among blacks. For example, a decomposition of a gap of five years may find that two of the five years is attributable to a higher rate of mortality due to CVD among blacks compared to whites. If the decomposition produces a negative estimate this implies that whites have a higher rate of CVD mortality for injuries compared to blacks. Thus, for each state, the contribution of each of the six causes to the black-white life expectancy gap was estimated for each year, producing forty trends (thirty-nine states and DC) and their credible intervals.

We also calculated state-level trends in age-standardized cause-specific mortality rates separately for each combination of race and gender, to aid the interpretation of the decomposition results. We used the US standard population from the year 2000 (National Cancer Institute, 2017b). The calculations described in this section were performed using functions written by the first author.

@@

A replication data set including the raw data and statistical code to reproduce the analysis and manuscript is publicly available on GitHub: https://github.com/corinne-riddell/BlackWhiteMortalityGap/. An online web application interactively hosts the findings (https://corinne-riddell.shinyapps.io/black-white-life-expectancy/).
